Community-Conditioned Language Models

### 2021-10-07 paper/paper1.tex
short paper, with CCLM embeddings

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-stratified-box}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are \subreddit{counting}, \subreddit{friendsafari}, and \subreddit{Fireteams}, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly structured and regular formats, which 
  are presumably conventional to these communities.}
%
The variation in information gain across communities is not 
related to perplexity, however (\cref{fig:ppl-info-gain}).

\begin{figure}
\begin{tikzpicture}
\begin{axis}[%
    axis x line*=bottom, axis y line*=left,
    xlabel=$\Ppl_{M_j}$, ylabel=$\IG_{M_j}$,
    width=7cm,
  ]
  \addplot+[%
    only marks,
    mark=*,
    mark options={scale=0.35, solid, color=violet!80, fill=violet},
    font=\tiny,
  ] table [x={lstm-3-ppl}, y={lstm-3-3-ig}] {floats/comm.csv};
\end{axis}
\end{tikzpicture}
\caption{
  Average community perplexity (x-axis) versus information gain (y-axis)
  for the best LSTM model ($l_c = 1$). Pearson's r=$-0.61$ (p<0.001).
}
\label{fig:ppl-info-gain}
\end{figure}

@@

\section{Latent Bayesian community classifiers} \label{sec:lmcc}
The perplexity of a CCLM gives us a measure of linguistic complexity
but it is not a measure of how linguistically \emph{specific} the
community is. However, the CCLM induces a latent Bayesian classifier,
which we call a language model-based community classifier
(LMCC).

So far, we have only loaded the community embedding which corresponds
to the actual community that the message originates from. By
evaluating the model's loss (and thus probability) for each potential community,
we can apply Bayesian reasoning to extract a latent community classifier for each message.
The probability that a given message $m$
belongs to a community $c_i$ is calculated as follows:
\[P(c=c_i \mid m) = P(m \mid c_i)\frac {P(m)} {P(c=c_i)}\]
In the above,
$P(c=c_i)$ is the frequency of community $c_i$ in the dataset, and
$P(m \mid c_i)$ (the probability of a message $m$
in the context of community $c_i$) is estimated by the CCLM. 
$P(m)$ is the absolute
probability of a message, which can be computed as
\[P(m) = \sum_i P(c=c_i) P(m\mid c=c_i ). \]
Since the test set is balanced across communities,
we have an even prior for $P(c=c_i)$, giving us
\[P(c=c_i\mid m) \propto  P(m\mid c_i).\]

We can characterise the performance of the LMCC with a confusion matrix
$C$, defined by $C_{ij} = P(c=c_i \mid m)$ for a given (test) set of messages $m$ 
from community $c_j$.
\[C_{ij} = \avg_{m\in M_j}P(c=c_i \mid m)\]
We observe that the diagonal is clearly above the rest: the LMCC generally
picks assigns the highest probability to the community where a message originates (\cref{fig:confusion}).

\begin{figure}
  \includegraphics[width=\linewidth]{floats/confusion_lstm-3-1}
\caption{
  LMCC confusion matrix for the LSTM model with $l_c=1$, showing $\log(P(c_j\mid c_i))$, 
  with $c_i$ on the x-axis and $c_j$ on the y-axis.
  Communities are sorted by indiscernability.
  (The first few communities predominantly use a language other than English, and are thus very discernable.)}
\label{fig:confusion}
\end{figure}

But we want a more quantitative measure, corresponding to the ability of the LMCC to correctly
recognise messages coming from a given community $c_j$. We could use
something like an F-score, but we prefer to use a perplexity-like metric,
which is has stronger information-theoretic grounds.
To this end, we define the \emph{linguistic indiscernibility} of community $c_j$, 
noted $\Ind_j$. 
First, note that each row of $C$ is a probability distribution,
since it is computed as the average of $P(c=c_i \mid m)$ for $m\in M_j$.
The exponential of the entropy of the $j$th row of \(C\) is therefore a perplexity score,
whose maximum is the total number of communities (510 for us).
Thus we define $\Ind_j$ by scaling this perplexity by the number of communities:
\[\Ind_j = \frac{e^{-\sum_i C_{ij} log(C_{ij})}}{|C_j|}\]
This way $Ind_j=0$ means that the LMCC is completely certain of its
choice, while $Ind_j=1$ means that equal probability is given to all
predictions.  In general $\Ind_j$ is a score of uncertainty of the
LMCC, given messages from a community $c_j$.

Indiscernibility of a given community is a robust measure: it is
almost perfectly correlated across CCLMs (such as transformers, LSTMs
and various values of $l_c$).  In fact, we observe a correlation
between the indiscernibility measures by every pair of models with
Pearson's $r \geq 0.96$ and $p < 0.0001$.

However, we observe that community-wise perplexity is only weakly (if at all) 
correlated with community indiscernibility under the corresponding latent 
LMCC (see \cref{tab:ind-corrs} and, for example, \cref{fig:cclm-lmcc-ppl}).
In other words, the linguistic complexity of a given community
is a poor predictor of how distinctive its language is.
%\jp{Conclusion: perplexity (linguistic complexity) is not correlated
  %to how recognisable a community is. Presumably (in general) some very specific
  %lexical items are characteristic of a community.}
%
\begin{table}
  \small
  \centering
  \begin{tabular}{llrrrr}
  \toprule
      & $l_c$ & $\Ppl_{M_j}$ & $\IG_{M_j}$ \\
  \midrule
  \multirow{4}{*}{LSTM}         & 0 &         0.21 &         -0.41 \\
                                & 1 &         0.19 &         -0.40 \\
                                & 2 &         0.19 &         -0.43 \\
                                & 3 &         0.06 &         -0.45 \\
  \midrule
  \multirow{4}{*}{Transformer}  & 0 &         0.19 &         -0.41 \\
                                & 1 &         0.23 &         -0.43 \\
                                & 2 &         0.16 &         -0.39 \\
                                & 3 &         0.12 &         -0.40 \\
  \bottomrule
  \end{tabular}
  \caption{ Pearson's $r$ correlation coefficient between community
    indiscernibility ($\Ind_j$) and two different predictors: CCLM
    perplexity on messages from community $c_j$ ($\Ppl_{M_j}$), and
    CCLM information gain on messages from $c_j$ ($\IG_{M_j}$). $p < 0.001$
    for all correlations except $\Ppl_{M_j}$ for the LSTM with 
  $l_c=3$ ($p=0.218$) and Transfomer with $l_c=3$ ($p=0.010$).}
  \label{tab:ind-corrs}
\end{table}

\begin{figure*}
  \begin{tikzpicture}
  \begin{groupplot}[
    group style={group size=2 by 1},
    legend style={anchor=north, legend columns=1, font=\tiny},
    every node near coord/.append style={font=\sffamily\tiny},%
    axis x line*=bottom, axis y line*=left,
  ]
  \nextgroupplot[xlabel=$\Ppl_{M_j}$, ylabel=$\Ind_j$]
      \addplot+[%
        only marks,
        mark=*,
        mark options={scale=0.35, solid, color=violet!80, fill=violet},
        font=\tiny,
      ] table [x={lstm-3-1-ppl}, y={lstm-3-1-indisc}] {floats/comm.csv};
  \nextgroupplot[xlabel=$\IG_{M_j}$, ylabel=$\Ind_j$, xmax=1.6]
      \addplot+[%
        only marks,
        mark=*,
        mark options={scale=0.35, solid, color=violet!80, fill=violet},
        font=\tiny,
      ] table [x={lstm-3-1-ig}, y={lstm-3-1-indisc}] {floats/comm.csv};
  \end{groupplot}
  \end{tikzpicture}
  \caption{%
    Community indiscernibility (y-axis), 
    in relationship to linguistic complexity (left) and 
    CCLM information gain (right). Shown here for the LSTM
    model with $l_c=1$. See \cref{tab:ind-corrs} for 
    Pearson's r correlations for all models.
    (Communities with \(\IG > 1.6\) (3 outliers) are hidden to make the correlation easier to observe.)
  }
  \label{fig:cclm-lmcc-ppl}
\end{figure*}

% We could also check if communities are well-separated in the SNAP community encodings

On the other hand, community indiscernibility is correlated with
the information gain of the CCLM
on messages from that community's test set
(see \cref{tab:ind-corrs} and, for example, \cref{fig:cclm-lmcc-ppl}).
Communities for which the language model benefits most from 
the conditioned model are generally easier for the LMCC to classify.
Intuitively this makes sense:
the more distinctive a community's language is,
the easier it is to guess where messages from that community originated
and the easier it is to predict the content of the message if you know the community.

Finally, we note that, even though community conditioning affects
perplexity very little overall, it provides large benefits in term of
the underlying Bayesian classifier. Indiscernibility is on average
\num{0.57} with a standard deviation of \num{0.22}. This means
that, while the linguistic variations do not allow our model to
pinpoint the originating community of a message, it can on average
eliminate 43\% of communities as possible origins 
for it. Furthermore, the information gain provided by
community-conditioning is a good predictor of how discernable the
community is.

% Finally, we investigate the correlation between the stability of a
% community and the perplexity of the LMCC for it. Stability is defined as TODO.

% - compute pearson correlation coefficient

% We see a positive correlation: this suggests that the more stable a
% community is, the more difficult it is to identify message as coming
% from it. What this suggests is that stable communities tend to use a
% varied and standard, subset of English, while unstable communities
% resort to more formulaic language. This may suggests that to 'fit' in an
% unstable community, one must use more obvious linguistic cues than in
% a stable community. In fact, it is reasonable to think that a new
% member will ostensibly use community-specifc language, but once
% established in a community, a member would tend not to do so.

>>>>

The conditioned models also perform differently across different communities---%
even among the best models, some communities have $\IG_{M_j} < 1$,
meaning that the CCLM performs worse than the unconditioned baseline for 
messages from that community.
For other communities $\IG_{M_j}$ is much higher, meaning that the CCLM performs better  
(\cref{fig:comm-stratified-box}).\footnote{
  Some of the communities with consistently high $IG_{M_j}$ 
  across all models are primarily non-English, 
  but surprisingly, not the three most extreme
  outliers. There are \subreddit{counting}, \subreddit{friendsafari}, and \subreddit{Fireteams}, 
  the later two of which are places where people coordinate to play video games
  together. The coordination has highly structured and regular formats, which 
  are presumably conventional to these communities.}
%
%The variation in information gain across communities is not 
%related to perplexity, however (\cref{fig:ppl-info-gain}).

@@
